Modelling the emergence of whisker barrels

### 2019-12-16 paper/briefpaper_stuedits/brief.tex
WIP on figure caption

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{k=1}^{N} c_{k}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one, connections decay at rate $\alpha$. Otherwise the connection density increases non-linearly ($k>1$) with the density of axon branching. Axon branching is modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux' of axonal branching. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

@@

\caption{\textbf{A.} Barreloid centres taken from \cite{haidarliu_size_2001} are shown as black circles. Barreloids are labelled according to the usual convention. Labels R1 to R5 are the rhinal whisker barreloids. The dashed grey line represents the boundary of the barreloid field; the dot-dash line shows the anterior-posterior axis. It is assumed that there are two linearly increasing, orthogonal signalling gradients (molecules T1 \& T2) in the thalamus (side graphs), such that the location of each barreloid within the graded thalamic signal field can differentiate the efferent axons, with the result that axons from each barreloid interact to varying degrees with cortical signalling gradients (shown in panel B). Coloured arrows show, for each barreloid, the magnitude (and sign) of the interaction that its thalamocortical projection will have with the signalling gradients in the cortex. The mappings from barreloid location to interaction strength are linear. \textbf{B.} Guidance molecules in the simulated cortex. Molecule 1 is expressed high to low in an approximately anterior to posterior gradient. Molecule 2 is expressed lateral to medial (analogously to ephrin - Vanderhaegen et al) \textbf{C.}  Whisker barrels from \cite{shimogori_fibroblast_2005}, with annotations for barrel identity and approximate axes.  \textbf{D-F.} Barrel cortex simulation. The maps indicate which thalomocortical axons make maximum connection density. Each hex element is coloured one of 52 unique colours depending on which thalamocortical axon makes the most connections in that element. The mapping between barrel identity and colour is given in F.  \textbf{G.} The Honda Dirichlet measure as a function of simluation time for the simulations show in D-F. \textbf{H.}  Guidance molecules in the simulated Fgf8 misexpression experiment. Molecule 1 is mirrored over the anterior-posterior axis \textbf{I.} The simulation indicates that the mirroring of the barrel field can occur as a result of the mirroring of guidance molecule 1.}
\label{fig:main}
\end{figure*}

>>>>

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one, connections decay at rate $\alpha$. Otherwise the connection density increases non-linearly ($k>1$) with the density of axon branching. Axon branching is modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux' of axonal branching. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

@@

\caption{\textbf{C.} Cortical whisker barrels from \cite{zheng_signal_2001}
  with annotations for approximate axes and lines drawn to join the
  interstices into a putative Voronoi diagram.  \textbf{B.}  Top: Stochastic
  initial condition for $a_i$ with $i=13$ (barrel C4). The roll-off towards
  the boundary (thin line) ensures the boundary conditions are
  satisfied. Bottom: Guidance molecules in the simulated cortex. Molecule 1 is
  expressed high to low in an approximately anterior to posterior
  gradient. Molecule 2 is expressed lateral to medial (analogously to ephrin -
  Vanderhaegen et al) \textbf{C.}  Barreloid centres for mystacial whiskers
  taken from \cite{haidarliu_size_2001} are shown as black circles. The dashed
  grey line shows the boundary of the barreloid field. It is assumed that
  there are two linearly increasing, orthogonal signalling molecules in the
  thalamus, such that the location of each barreloid within the graded
  thalamic signal field can differentiate the efferent axons, with the result
  that axons from each barreloid interact to varying degrees with the cortical
  signalling gradients shown in B. Coloured arrows show, for each barreloid,
  the magnitude (and sign) of the interaction that its thalamocortical
  projection will have with each signalling gradient in the cortex. The
  mappings from barreloid location to interaction strength are
  linear. \textbf{D.}  Barrel cortex simulation. The maps indicate which
  thalomocortical axons make maximum connection density. Each hex element is
  coloured one of 41 \cmnt{FIXME: Got to here} unique colours depending on which thalamocortical axon
  makes the most connections in that element. The mapping between barrel
  identity and colour is given in F.  \textbf{G.} The Honda Dirichlet measure
  as a function of simluation time for the simulations show in
  D-F. \textbf{H.}  Guidance molecules in the simulated Fgf8 misexpression
  experiment. Molecule 1 is mirrored over the anterior-posterior axis
  \textbf{I.} The simulation indicates that the mirroring of the barrel field
  can occur as a result of the mirroring of guidance molecule 1.}
\label{fig:main}
\end{figure*}
